Patterns of stress tolerance versus resistance in genome-wide expression data

### 2014-07-16 doc/ApTranscriptome_manuscript.Rmd
Updates to ms

The Trinity *de novo* transcriptome assembly included 126,172 genes with a total length of 100 million bp. Filtering to remove redundant or chimeric reads resulted in an assembly with 105,536 genes. The total length was 63 million bp with an N~50~ length of 895 bp and a mean gene size of 593 bp. Of the 105,536 filtered genes, 55,432 had hits to the NCBI-nr database. Of these, 38,711 genes mapped to GO terms, 1,659 genes were identified to an enzyme and 18,935 genes mapped to a domain with >50% coverage. DeconSeq identified 5,675 genes as known contaminants from human, virus, bacterial or archael sources, leaving 99,861 clean transcripts. Of these, 47% have best hits to genes from the class Insecta [(Supplementary Krona plot)](http://johnstantongeddes.org/assets/files/ApTranscriptome_FA.html).

```{r mosaic_plot, echo=FALSE, results='hide'}
tiff("mosaic_plot.tiff")
mosaic.colors <- brewer.pal(5, "Blues")
par(mar = c(5.1, 3.1, 2.1, 7.5))
mosaicplot(tt2, ylab = expression(italic("A. picea")), xlab = "", main = "", color = mosaic.colors, cex.axis = 0.01)
labels <- c("Low", "Intermediate", "High", "Bimodal", "NotResp")
text(c(.1,.5,.75,.85,.99), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE, font = 1)
text(.5, par("usr")[4]+.1, "A. carolinensis", xpd = TRUE, font = 3)
text(.15, .7, "2,594")
text(.85, .03, "223", cex = .75)
legend(1.05, .9, fill = mosaic.colors, c("Low", "Intermediate", "High", "Bimodal", "NotResp"), xpd=TRUE)
dev.off()
```

![Mosaic plot of the number of thermally-responsive genes by expression type. The width of the columns is proportional to the number of genes for each expression type for *A. carolinensis* while the height of the rows is proportional to the number of genes for each expression type for *A. picea.* The rows are colored by expression type for *A. picea*. The number of genes are labeled in two boxes for scale.](mosaic_plot.tiff)

We identified `r prettyNum(length(which(RxNpval$pval < 0.05)), big.mark = ",", scientific = FALSE)` genes with overall model fits of *P* < 0.05, retaining `r prettyNum(length(which(RxNpval$padj < 0.05)), big.mark = ",", scientific = FALSE)` after adjusting *P* values for false discovery rate. Of these genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` significantly change expression with temperature including `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` that have a species x temperature interaction. The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not thermally-responsive in one species (*NotResp*) differed significantly between the species ($\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Table 1). The density function for the temperature of maximum expression (Fig. 1) showed that the majority of responsive genes were in the *Low* category for both species, with *A. picea* having more *Low* genes than *A. carolinensis*. More genes had *Intermediate* expression in *A. carolinensis* than in *A. picea*, while the two species had a similar number of *High* genes (Fig. 1, Table 1).

The reaction norms of thermally-responsive genes between species differed by mean expression (*t*~`r t.mean$parameter`~ = `r t.mean$statistic`, *P* < `r max(t.mean$p.value, 0.001)`) which was about `r round(exp(t.mean$estimate))` TPM greater in *A. carolinensis* than *A picea*, and mean slope (*t*~`r t.slope$parameter`~ = `r t.slope$statistic`, *P* < `r max(t.slope$p.value, 0.001)`), which was about `r round(exp(t.mean$estimate))` $\Delta$TPM/degree greater in *A. picea*. Curvature (*t*~`r t.curvature$parameter`~ = `r t.curvature$statistic`, *P* < `r round(max(t.curvature$p.value, 0.001), 3)`) and wiggle (*t*~`r t.wiggle$parameter`~ = `r t.wiggle$statistic`, *P* < `r max(t.wiggle$p.value, 0.001)`) did not differ between species. Of the total differences ($\Delta T$) in reaction norms between species for each gene, about `r round(qlmean[2] * 100, 0)`% (95% CI `r paste(round(qlmean[1] * 100, 0), "--", round(qlmean[3] * 100, 0))`%) and `r round(qlslope[2] * 100, 0)`% (95% CI `r paste(round(qlslope[1] * 100, 0), "--", round(qlslope[3] * 100, 0))`%) were due to shifts in mean and slope, respectively.

We predicted that we would observe that genes up-regulated in response to thermal stress would have increased basal expression in the other species. Instead, we found that for genes expressed at *High* temperatures in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`), genes expressed at *Low* temperatures in *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`) and *Intermediate* genes in both (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`), basal expression near optimum temperatures of 19.5°C was always higher in *A. carolinensis*.

Overall greater thermal tolerance of one species would be indicated by finding that gene expression is stable across a greater portion of the thermal gradient. For *Bimodal* expressed genes, we found no difference (*t*~df=1406~ = 0.31, *P* = 0.76) between species in the variance of the reaction norms for thermally-responsive genes. In contrast, *Intermediate* genes from *A. carolinensis* had a broader thermal range of stable gene expression than those from *A. picea* (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64). In addition, the species differed in the temperatures at which thermally-responsive genes were activated. For genes with increased expression at *High* temperatures, we found that *A. carolinensis* genes were activated on average at 33.8°C compared to 32.9°C for *A. picea* (Fig. 2). For genes with increased expression at *Low* temperatures, *A. carolinensis* genes were activated on average at 12.8°C compared to 13°C for a *A. picea* (Fig. 2).

```{r T_on_plot, echo=FALSE, warning = FALSE, fig.cap = "Histogram with smooth density estimate of temperature of maximum change in expression for genes that have increased expression at *Low* and *High* temperatures. Bars are Probability density function of peak expression for all thermally-responsive in *A. carolinensis* (blue) and *A. carolinensis* (red) species"}
T_on_plot <- ggplot(Ap.T_on, aes(x = T_on, y = ..density.., colour = colony, group = colony, fill = colony)) +
    geom_histogram(position = "dodge", binwidth = 3) +
    geom_density(alpha = 0.2, adjust = 3) + 
    scale_y_continuous(name="Density") +
    scale_x_continuous(name=expression("Temperature of gene activation"))
print(T_on_plot)
```

### Molecular comparisons

As the molecular pathways involved in thermal response to cold and hot temperatures differ, we performed gene set enrichment analysis on each expression category separately. Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "High"), ])` enriched (*P* < 0.01) for *High* transcripts, including categories such as `r Ap.gsea1[which(Ap.gsea1$Type == "High"), "Term"][1]` and `r Ap.gsea1[which(Ap.gsea1$Type == "High"), "Term"][6]` (Table S1). For *Low* genes, `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "Low"), ])` GO terms were enriched including some implicated in stress response (e.g. `r Ap.gsea1[which(Ap.gsea1$Type == "Low"), "Term"][18]`) and cold tolerance (e.g. `r Ap.gsea1[which(Ap.gsea1$Type == "Low"), "Term"][6]`) (Table S1). For *Bimodal* genes, `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "Bimodal"), ])` GO terms were enriched including `r length(grep("regulation", Ap.gsea1[which(Ap.gsea1$Type == "Bimodal"), "Term"]))` *regulation* terms (e.g. regulation of gene expression) indicating that these are general genes involved in activating responses to change (Table S1). Finally, for *Intermediate* genes `r nrow(Ap.gsea1[which(Ap.gsea1$Type == "Intermediate"), ])` GO terms were enriched, primarily metabolic or biosynthetic processes (Table S1).

@@

```{r rxn_plot, echo = FALSE, warning = FALSE, fig.cap = "Gene expression in transcripts per million (TPM) for thermally-reactive genes in the GO category *response to stress* for each colony"}
rxnplot <- ggplot(resp.TPM.dt.sub[GO0006950.responsive$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) +
  facet_grid(. ~ colony2) +
  scale_y_continuous(name = "Expression (scaled)") +
  scale_x_continuous(name = expression(paste("Temperature", degree, "C")))
print(rxnplot)
```

>>>>

### Transcriptome assembly

The Trinity *de novo* transcriptome assembly included 126,172 genes with a total length of 100 million bp. Filtering to remove redundant or chimeric reads resulted in an assembly with 105,536 genes. The total length was 63 million bp with an N~50~ length of 895 bp and a mean gene size of 593 bp. Of the 105,536 filtered genes, 55,432 had hits to the NCBI-nr database. Of these, 38,711 genes mapped to GO terms, 1,659 genes were identified to an enzyme and 18,935 genes mapped to a domain with >50% coverage. We removed 5,675 genes identified as known contaminants, leaving 99,861 clean transcripts. Of these, 47% have best hits to genes from the class Insecta [(Supplementary Krona plot)](http://johnstantongeddes.org/assets/files/ApTranscriptome_FA.html).

```{r mosaic_plot, echo=FALSE, results='asis', fig.cap = "Mosaic plot of the number of thermally-responsive genes by expression type. The width of the columns is proportional to the number of genes for each expression type for *A. carolinensis* while the height of the rows is proportional to the number of genes for each expression type for *A. picea.* The rows are colored by expression type for *A. picea*. The number of genes are labeled in two boxes for scale."}
mosaic.colors <- brewer.pal(5, "Blues")
par(mar = c(5.1, 3.1, 2.1, 7.5))
mosaicplot(tt2, ylab = expression(italic("A. picea")), xlab = "", main = "", color = mosaic.colors, cex.axis = 0.01)
labels <- c("Low", "Intermediate", "High", "Bimodal", "NotResp")
text(c(.1,.5,.75,.85,.99), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE, font = 1)
text(.5, par("usr")[4]+.1, "A. carolinensis", xpd = TRUE, font = 3)
text(.15, .7, "2,594")
text(.85, .03, "223", cex = .75)
legend(1.05, .9, fill = mosaic.colors, c("Low", "Intermediate", "High", "Bimodal", "NotResp"), xpd=TRUE)
```

We identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` genes with expression patterns that are thermally-responsive at 5% FDR. Of these, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` have an expression pattern that differs between the two species. The number of transcripts in each expression category differed significantly between the species ($\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig. 1). In both species, the majority of genes were in the *Low* category (Fig. 1). For *A. carolinensis*, this was followed by *Intermediate*, *High* and then *Bimodal*, whereas for *A. picea* the second largest category was *Bimodal* followed by *High* and finally *Intermediate* (Fig. 1). There were `r colSums(tt2)[which(colnames(tt2) == "NotResp")]` genes that were responsive in *A. carolinensis* but not *A. picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` that were responsive in *A. picea* but not *A. carolinensis*.

### Species comparisons

To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the temperature of gene activation for *High* and *Low* genes, and the variance in expression for the *Intermediate* and *Bimodal* genes. The *High* genes in *A. carolinensis* were activated on average at `r T_on.high.ttest$estimate[1]`°C compared to `r T_on.high.ttest$estimate[2]`°C for *A. picea* (*t*~df=`r T_on.high.ttest$parameter`~ = `r T_on.high.ttest$statistic`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For genes with increased expression at *Low* temperatures, there was no difference in the mean temperature of gene activation (*t*~df=`r T_on.low.ttest$parameter`~ = `r T_on.low.ttest$statistic`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2). For *Bimodal* expressed genes, we found no difference between species in the variance of the reaction norms (*t*~df=`r round(t.varbim$parameter, 0)`~ = `r round(t.varbim$statistic, 2)`, *P* < `r round(t.varbim$p.value, 2)`). In contrast, *Intermediate* genes from *A. carolinensis* had a broader thermal range of stable gene expression than those from *A. picea* (*t*~df=`r round(t.varint$parameter, 0)`~ = `r round(t.varint$statistic, 2)`, *P* < `r max(t.varint$p.value, 0.001)`, 95% CI difference `r paste(round(t.varint$conf.int[1], 3), "--", round(t.varint$conf.int[2], 3))`).

```{r T_on_plot, echo=FALSE, warning = FALSE, fig.cap = "Histogram with smooth density estimate of temperature of maximum change in expression for genes that have increased expression at *Low* and *High* temperatures. Bars are Probability density function of peak expression for all thermally-responsive in *A. carolinensis* (blue) and *A. carolinensis* (red) species"}

T_on_plot <- ggplot(Ap.T_on, aes(x = T_on, y = ..density.., colour = Species, group = Species, fill = Species)) +
    facet_grid(. ~ type) +
    geom_histogram(position = "dodge", binwidth = 3) +
    geom_density(alpha = 0.2, adjust = 3) + 
    scale_y_continuous(name="Density") +
    scale_x_continuous(name=expression("Temperature of gene activation"))
print(T_on_plot)
```

We looked for evidence of genetic assimilation by comparing gene expression at intermediate temperatures for genes that were up-regulated at low temperatures in *A. carolinensis* or high temperatures in *A. picea*. Contrary to expectations, for *Low* genes in *A. carolinensis* gene expression at intermediate temperatures was lower in *A. picea* than *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`) . Moreover, for *High* genes in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`) and *Intermediate* genes in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`), gene expression at intermediate temperatures was always higher in *A. carolinensis*.

In addition to shifts in position, we quantified the differences between species in the mean and shape of the reaction norms of thermally-responsive genes. Species differed by mean expression (*t*~df=`r t.mean$parameter`~ = `r t.mean$statistic`, *P* < `r max(t.mean$p.value, 0.001)`) which was about `r round(exp(t.mean$estimate))` TPM greater in *A. carolinensis* than *A picea*, and mean slope (*t*~df=`r t.slope$parameter`~ = `r t.slope$statistic`, *P* < `r max(t.slope$p.value, 0.001)`), which was about `r round(exp(t.mean$estimate))` $\Delta$TPM/degree greater in *A. picea*. Curvature (*t*~df=`r t.curvature$parameter`~ = `r t.curvature$statistic`, *P* < `r round(max(t.curvature$p.value, 0.001), 3)`) and wiggle (*t*~df=`r t.wiggle$parameter`~ = `r t.wiggle$statistic`, *P* < `r max(t.wiggle$p.value, 0.001)`) did not differ between species. Of the total differences ($\Delta T$) in reaction norms between species for each gene, about `r round(qlmean[2] * 100, 0)`% (95% CI `r paste(round(qlmean[1] * 100, 0), "--", round(qlmean[3] * 100, 0))`%) and `r round(qlslope[2] * 100, 0)`% (95% CI `r paste(round(qlslope[1] * 100, 0), "--", round(qlslope[3] * 100, 0))`%) were due to shifts in mean and slope, respectively.

### Gene set enrichment analysis

As the molecular pathways involved in thermal response to cold and hot temperatures differ, we performed gene set enrichment analysis on each expression category separately. Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched (*P* < 0.01) terms for *High* transcripts. Clustering of terms by semantic similarity yielded 3 clusters, with "regulation of biosynthetic/metabolic process" being the largest group.

For *Low* genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` GO terms were enriched. These terms grouped into 6 clusters, including "protein metabolic process", "lipid metabolic process", and "".

including some implicated in stress response (e.g. `r Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), "Term"][18]`) and cold tolerance (e.g. `r Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), "Term"][6]`) (Table S1). For *Bimodal* genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched including `r length(grep("regulation", Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), "Term"]))` *regulation* terms (e.g. regulation of gene expression) indicating that these are general genes involved in activating responses to change (Table S1). Finally, for *Intermediate* genes `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched, primarily metabolic or biosynthetic processes (Table S1).

@@
